---
title: "Untargeted characterization of Amazonian barks extracts plants by Liquid and Gas Chromatography - Mass Spectrometry | Liquid Chromatography - Mass Spectrometry (LC-MS) data analysis 1"
author: "Jefferson Pastuña"
date: "2023-08-31"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

## Introduction

This R Script aims to record the procedure given for metabolic profiling of 4 species of plant used in Amazonia folk medicine. Each step has a brief explanation, code and graphics.

The workflow used was taken from ["notame": Workflow for Non-Targeted LC–MS Metabolic Profiling](https://doi.org/10.3390/metabo10040135). Which offers a wide variety of functions to perform metabolomic profile analysis.

## Before to start

The "notame" package accepts as input a feature table that can be obtained through software such as MZmine, MS-DIAL, among others. In this case, the feature table was obtained with the help of MZmine. The (*.csv) file was fixed to obtain the final feature table input.

## Notame workflow

As a first step for the analysis,"notame" package and other dependency packages were installed.

```{r echo=TRUE, message=FALSE}

# Notame package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("antonvsdata/notame")
# Notame library call
library(notame)
# Dependency packages installation
install_dependencies

```

Then, a main path and a log system was added to have a record of each process executed.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Main path
ppath <- "../Arajuno_Stem_Bark/"
# Log system
init_log(log_file = paste0(ppath, "Result/LCMS/Firts_LCMS_Dataset_log.txt"))

```

Next, the MZmine suitable feature list was imported.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

data1 <- read_from_excel(file = "Data/Firts_LCMS_Dataset_to_R.xlsx", sheet = 5,
                         corner_row = 5, corner_column = "J",
                         split_by = c("Column", "Ion Mode"))

```

Once the data is read, the next step was to create a MetaboSet in order to obtain a specific R object.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

modes1 <- construct_metabosets(exprs = data1$exprs,
                               pheno_data = data1$pheno_data,
                               feature_data = data1$feature_data,
                               group_col = "Group")

```

We can visualize the raw data in order to inspect the processing routines.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Data extraction
mode1_test <- modes1$RP_NEG
# Boxplot of raw data
raw_bp1 <- plot_sample_boxplots(mode1_test,
                                order_by = "Group",
                                fill_by = "Group")
# PCA of raw data
raw_pca1 <- plot_pca(mode1_test,
                     center = TRUE,
                     shape = "Group",
                     color = "Group")
# Package to plots visualization in a same windows
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("thomasp85/patchwork")
library(patchwork)
# Plot
raw_pca1 + raw_bp1

```

## Preprocessing

The first step of the preprocessing is to change the features with value equal to 0 to NA.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Data extraction
mode1 <- modes1$RP_NEG
# Change 0 value to NA
mode1 <- mark_nas(mode1, value = 0)

```

Then, features with low detection rate are first flagged and then could be removed. The "notame" package employs two criteria to select this features. First, is the feature presence in a percentage of QC injections, and then the feature presence in a percentage within a sample group or class.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Low detection rate
mode1 <- flag_detection(mode1, qc_limit = 3/4, group_limit = 0)

```

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Some statistics after low detection algorithm
#visualizations(mode1, prefix = paste0(ppath, "Figure/LCMS/", "Low_Detection"))

```

With these values, features which that were not detected in 75% of the QC injections will be flagged as low detection rate.

The next step for preprocessing correspond to drift correction. The drift correction can be applied by smoothed cubic spline regression.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drift correction
corrected1 <- correct_drift(mode1)
# Flag low quality features
corrected1 <- flag_quality(corrected1)

```

Then we can visualize the data after drift correction.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
corr_bp1 <- plot_sample_boxplots(corrected1,
                                 order_by = "Group",
                                 fill_by = "Group")
# PCA
corr_pca1 <- plot_pca(corrected1,
                      center = TRUE,
                      shape = "Group",
                      color = "Group") 
# Plot
corr_pca1 + corr_bp1

```

Contaminant peaks based on the process blank will be removed.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Removal of contaminants
corrected1_no_blank <- flag_contaminants(corrected1,
                                         blank_col = "Group",
                                         blank_label = "PB",
                                         flag_thresh = 0.8,
                                         flag_label = "Contaminant")
# Removal blank group from dataset
corrected1_no_blank <- corrected1_no_blank[, corrected1_no_blank$Group != "PB"]

```

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Some statistics after feature blank detection algorithm
#visualizations(corrected1_no_blank, prefix = paste0(ppath, "Figure/LCMS/", "No_Blank"))

```

The next step is feature clustering. This step helps us reduce the number of features of the same molecule that were split due to ionization behavior (In-source fragmentation for example).

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

clustered1 <- cluster_features(corrected1_no_blank,
                               rt_window = 1/60,
                               all_features = TRUE,
                               corr_thresh = 0.90,
                               d_thresh = 0.8
                               #plotting = TRUE,
                               #prefix = paste0(ppath, "Cluster/LCMS/LCMS_Cluster")
                               )
compressed1 <- compress_clusters(clustered1)

```

We can inspect data after clustering algorithm.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
compr_bp1 <- plot_sample_boxplots(compressed1,
                                  order_by = "Group",
                                  fill_by = "Group")
# PCA
compr_pca1 <- plot_pca(compressed1,
                       center = TRUE,
                       shape = "Group",
                       color = "Group") 
# Plot
compr_pca1 + compr_pca1

```

The next step imputes the data.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# To clean data
set.seed(123)
imputed1 <- impute_rf(compressed1)

```

We can inspect PCA plot after imputation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
imp_bp1 <- plot_sample_boxplots(imputed1,
                                order_by = "Group",
                                fill_by = "Group")
# PCA
imp_pca1 <- plot_pca(imputed1,
                     center = TRUE,
                     shape = "Group",
                     color = "Group")
# Plot
imp_pca1 + imp_bp1

```

Finally, the data is ready to be exported and proceed with the statistical analysis.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

#save(imputed1, file = paste0(ppath, "Result/LCMS/First_LCMS_output.RData"))

```

# PCA plots

Droping flagged features

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Extract clean data
no_flag1 <- drop_flagged(imputed1)

# Extracting feature height table
peak_height1 <- exprs(no_flag1)
# Extracting samples information
pheno_data1 <- no_flag1@phenoData@data
# Extracting feature information
feat_data1 <- no_flag1@featureData@data

```

Preparing data and transposing feature table.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Transposing feature height table
transp_table1  <- t(peak_height1)
# Changing NA to 0 
transp_table1[is.na(transp_table1)]=0
# Centering and Scaling features
neg_pca1 <- prcomp(transp_table1, center = TRUE, scale. = TRUE)

```

Plotting PCA results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Library to left_join use
library(dplyr)
# PCA scores
scores1 <- neg_pca1$x %>%
  data.frame %>%                         # Convert to data frames
  mutate(Sample_ID = rownames(.)) %>%    # Create a new column with the sample names
  left_join(pheno_data1)                 # Adding metadata
# PCA plot
ggplot(scores1,
       aes(PC1, PC2, shape = Species, color = Species)) +
  geom_point(size = 3) +
  guides(x=guide_axis(title = "PC1 (29.03 %)"),
         y=guide_axis(title = "PC2 (18.43 %)")) +
  theme_classic()
# Save plot
#ggsave("Result/LCMS/Firts_LCMS_PCA.png", width = 5, height = 4, device='png', dpi="print")

```

Plotting loading results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

loadings1 <- neg_pca1$rotation %>%           # Extract loadings
  data.frame(Cluster_ID = rownames(.))       # New column with feature name

```

Creating an artificial table with feature ID and compound name column.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Exporting notame output to find and filter identified metabolites
#write_to_excel(clustered, "Result/LCMS/Firts_clustered_features.xlsx")
# Load a metabolite name table
metab_name1 <- readxl::read_excel("Data/Firts_LCMS_Metabolites.xlsx", 1)
# Creating a new small table of the annotated compounds
neg_compouds1 <- left_join(metab_name1, loadings1)
# Plotting results
figure_1b <- ggplot(loadings1, aes(PC1, PC2)) +
  geom_point(alpha = 0.2) +
  theme_classic() +
  geom_point(data = neg_compouds1,
             aes(shape = `Identification level`,
                 color = `Identification level`),
             size = 2) +
  ggrepel::geom_label_repel(data = neg_compouds1,
                            aes(label = Metabolite_name),
                            box.padding = 0.8,
                            label.padding = 0.3,
                            label.r = 0.3,
                            cex = 3) +
  guides(x=guide_axis(title = "PC1 (29.03 %)"),
         y=guide_axis(title = "PC2 (18.43 %)")) +
  ggsci::scale_color_aaas() +
  theme(legend.position = c(0.88, 0.50),
        legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray")
figure_1b
# Save plot
#ggsave('Result/LCMS/Firts_LCMS_Loadings_to_DQ.pdf', width = 8, height = 4, device='pdf', dpi="print")

```

# Heat map plot

ComplexHeatmap package and dependency installation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# ComplexHeatmap package installation
#if (!requireNamespace("BiocManager", quietly=TRUE))
#    install.packages("BiocManager")
#BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)

# ColorRamp2 package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("jokergoo/colorRamp2")
library(colorRamp2)

# Cowplot package installation
#install.packages("cowplot")
library(cowplot)

```

Extracting and loaded of identified metabolites abundance.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Exporting "notame" output to find and filter height of identified metabolites
#write_to_excel(no_flag, "Result/LCMS/Firts_no_flag_features.xlsx")
# Metabolite name table
metab_name_hm1 <- readxl::read_excel("Data/Firts_LCMS_Metabolite_HeatMap.xlsx", 2)
# Logarithmic scale
hm_scl1 <- metab_name_hm1[, 6:9] %>% as.matrix %>% log10()
rownames(hm_scl1) <- metab_name_hm1$Metabolite_name
# Metabolite classification
metab_class1 <- metab_name_hm1 %>% select(Class = Superclass, Metabolite = Metabolite_name)
# Metabolite class to HeatMap anotation
met_class_annotation1 <-  metab_class1 %>% select(Class) %>% 
  as.matrix()
rownames(met_class_annotation1) <- metab_class1$Metabolite
# Top information
top_info1 <- data.frame(Species = c(rep("Ar"),
                                    rep("Cg"),
                                    rep("Ml"),
                                    rep("Ps")))
rownames(top_info1) <- paste(top_info1$Species, rep(c(1, 2, 3, 4)))
top_info1 <- as.matrix(top_info1)

```

Scaling, row and top heatmap anotation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

set.seed(2024)
# Metabolite class color
cols_metclass1 <- c("Benzenoids" = "#800000FF",
                    "Organic oxygen compounds" = "#8A9045FF",
                    "Lipids and lipid-like molecules" = "#FFA319FF",
                    "Phenylpropanoids and polyketides" = "#FF5A5A",
                    "Organic acids and derivatives" = "#155F83FF")
# Add row anotation to HeatMap
hm_row_ann1 <- rowAnnotation(Metabolite = met_class_annotation1,
                             col = list(Metabolite = cols_metclass1),
                             show_annotation_name = T,
                             show_legend=F)
# Species color
cols_species1 <- c("Ar" = "#e6550d",
                   "Cg" = "#91cf60",
                   "Ml" = "#33d2ff",
                   "Ps" = "#d957f8")
# Add top anotation to HeatMap
top_info_ann1 <- HeatmapAnnotation(Species = top_info1,
                                   col = list(Species = cols_species1),
                                   show_annotation_name = T,
                                   show_legend=F,
                                   border = TRUE)
# Color scale
mycol1 <- colorRamp2(c(3, 5, 7),
                     c("blue", "white", "red"))
#split = sample(letters[1:3], 15, replace = T)
# Heatmap matrix plotting
hm_plot1 <- Heatmap(hm_scl1,
                    col = mycol1,
                    border_gp = grid::gpar(col = "black", lty = 0.05),
                    rect_gp = grid::gpar(col = "black", lwd = 0.75),
                    clustering_distance_columns = "euclidean",
                    clustering_method_columns = "complete",
                    top_annotation = top_info_ann1,
                    right_annotation = hm_row_ann1,
                    column_names_gp = gpar(fontface = "italic"),
                    show_heatmap_legend = F,
                    #row_split = split,
                    row_km = 3,
                    column_km = 2)
hm_plot1

```

Adding legends to heatmap.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Color scale legend
lgd1a <- Legend(col_fun = mycol1,
                title = "log10 abundance",
                #at = seq(7),
                direction = "horizontal" )
# Plants species legend
lgd2b <- Legend(labels = gt_render(c("*A. rigidum*",
                                     "*C. guianensis*",
                                     "*M. laevis*",
                                     "*P. sagotianum*")),
                legend_gp = gpar(fill = cols_species1),
                title = "Plant species", ncol = 1)
# Metabolite class Legend
lgd3c <- Legend(labels = c(unique(metab_class1$Class)),
                legend_gp = gpar(fill = cols_metclass1),
                title = "Metabolite superclass", ncol = 2)

```

ComplexHeatmap plot

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

set.seed(2024)
# Converting to ggplot
gg_heatmap1 <- grid.grabExpr(draw(hm_plot1))
gg_heatmap1 <- ggpubr::as_ggplot(gg_heatmap1)
# Legends
all_legends1 <- packLegend(lgd1a, lgd2b, lgd3c, direction = "horizontal")
gg_legend1 <- grid.grabExpr(draw(all_legends1))
gg_legend_fn1 <- ggpubr::as_ggplot(gg_legend1)
# Heatmap plot
figure_1c <- plot_grid(gg_legend_fn1,
                       gg_heatmap1, ncol = 1,
                       rel_heights = c(1.03697, 6.6))
figure_1c
# Save heatmap plot
#ggsave(filename = "Result/LCMS/Firts_LCMS_Heatmap_to_DQ2.pdf", plot = gcms_hm,
      #width = 5, height = 4, units = "in", dpi = 300, scale = 1.7)

```

# Exporting plot

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

library(gridExtra)

Figure_11 <- arrangeGrob(figure_1a,
                         figure_1b,
                         figure_1c,
                         layout_matrix = rbind(c(1, 2),
                                               c(3, 3)))

figure_oneone <- ggpubr::as_ggplot(Figure_11) +
  draw_plot_label(label = LETTERS[1:3],
                  x = c(0, 0.5, 0),
                  y = c(.99, .99, .475))
# (*.pdf) file
#ggsave(filename = "Result/LCMS/Figure_1.pdf", plot = figure_one,
#      width = 130, height = 130, units = "mm", dpi = 300, scale = 2.5)
# (*.png) file
#ggsave(filename = "Result/LCMS/Figure_1.png", plot = figure_one,
#      width = 130, height = 130, units = "mm", dpi = 300, scale = 2.5)
# (*.jpg) file
ggsave(filename = "Result/LCMS/Figure_1.jpg", plot = figure_oneone,
      width = 140, height = 130, units = "mm", dpi = 300, scale = 2.5)

```

Finish a record.

```{r}

finish_log()

```

