---
title: "Untargeted characterization of Amazonian barks extracts plants by Liquid and Gas Chromatography - Mass Spectrometry | Liquid Chromatography - Mass Spectrometry (LC-MS) data analysis 1"
author: "Jefferson Pastuña"
date: "2023-08-31"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

## Introduction

This R Script aims to record the procedure given for metabolic profiling of 4 species of plant used in Amazonia folk medicine. Each step has a brief explanation, code and graphics.

The workflow used was taken from ["notame": Workflow for Non-Targeted LC–MS Metabolic Profiling](https://doi.org/10.3390/metabo10040135). Which offers a wide variety of functions to perform metabolomic profile analysis.

## Before to start

The "notame" package accepts as input a feature table that can be obtained through software such as MZMine, MSDial, among others. In this case, the feature table was obtained with the help of MZmine. The (*.txt) file was fixed to obtain the final feature table input.

## Notame workflow

As a first step for the analysis,"notame" package and other dependency packages were installed.

```{r echo=TRUE, message=FALSE}

# Notame package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("antonvsdata/notame")
# Notame library call
library(notame)
# Dependency packages installation
install_dependencies

```

Then, a main path and a log system was added to have a record of each process executed.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Main path
ppath <- "F:/Jefferson-Pastuna/Arajuno_Stem_Bark/"
# Log system
init_log(log_file = paste0(ppath, "Result/LCMS/LCMS_log.txt"))

```

Next, the MZmine suitable feature list was imported.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

data <- read_from_excel(file = "Data/LCMS_to_R.xlsx", sheet = 5, 
                        corner_row = 5, corner_column = "J", 
                        split_by = c("Column", "Ion Mode"))

```

Once the data is read, the next step was to create a MetaboSet in order to obtain a specific R object.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

modes <- construct_metabosets(exprs = data$exprs, 
                              pheno_data = data$pheno_data, 
                              feature_data = data$feature_data,
                              group_col = "Group")

```

We can visualize the raw data in order to inspect the processing routines.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Data extraction
mode_test <- modes$RP_NEG
# Boxplot of raw data
raw_bp <- plot_sample_boxplots(mode_test,
                               order_by = "Group",
                               fill_by = "Group")
# PCA of raw data
raw_pca <- plot_pca(mode_test,
                       center = TRUE,
                       shape = "Group",
                       color = "Group")
# Package to plots visualization in a same windows
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("thomasp85/patchwork")
library(patchwork)
# Plot
raw_pca + raw_bp

```

## Preprocessing

The first step of the preprocessing is to change the features with value equal to 0 to NA.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Data extraction
mode <- modes$RP_NEG
# Change 0 value to NA
mode <- mark_nas(mode, value = 0)

```

Then, features with low detection rate are first flagged and then could be removed. The notame package employs two criteria to select this features. First, is the feature presence in a percentage of QC injections, and then the feature presence in a percentage within a sample group or class.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Low detection rate
mode <- flag_detection(mode, qc_limit = 3/4, group_limit = 0)

```

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Some statistics after low detection algorithm
visualizations(mode, prefix = paste0(ppath, "Figure/LCMS/", "Low_Detection"))

```

With these values, features which that were not detected in the 75% of the QC injections and 0% of sample groups will be flagged as low detection rate.

The next step for preprocessing correspond to drift correction. The drift correction can be applied by smoothed cubic spline regression.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drift correction
corrected <- correct_drift(mode)
# Flag low quality features
corrected <- flag_quality(corrected)

```

Then we can visualize the data after drift correction.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
corr_bp <- plot_sample_boxplots(corrected,
                                      order_by = "Group",
                                      fill_by = "Group")
# PCA
corr_pca <- plot_pca(corrected,
                        center = TRUE,
                        shape = "Group",
                        color = "Group") 
# Plot
corr_pca + corr_bp

```

Contaminant peaks based on the Process Blank will be removed.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Removal of contaminants
corrected_no_blank <- flag_contaminants(corrected,
                                        blank_col = "Group",
                                        blank_label = "PB",
                                        flag_thresh = 0.8,
                                        flag_label = "Contaminant")
# Removal blank group from dataset
corrected_no_blank <- corrected_no_blank[, corrected_no_blank$Group != "PB"]

```

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Some statistics after feature blank detection algorithm
visualizations(corrected_no_blank, prefix = paste0(ppath, "Figure/LCMS/", "No_Blank"))

```

The next step is feature clustering. This step helps us reduce the number of features of the same molecule that were split due to ionization behavior.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

clustered <- cluster_features(corrected_no_blank,
                              rt_window = 1/60,
                              all_features = TRUE,
                              corr_thresh = 0.90,
                              d_thresh = 0.8,
                              plotting = TRUE,
                              prefix = paste0(ppath, "Cluster/LCMS/LCMS_Cluster")
                              )
compressed <- compress_clusters(clustered)

```

We can inspect data after clustering algorithm.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
compr_bp <- plot_sample_boxplots(compressed,
                                 order_by = "Group",
                                 fill_by = "Group")
# PCA
compr_pca <- plot_pca(compressed,
                        center = TRUE,
                        shape = "Group",
                        color = "Group") 
# Plot
compr_pca + compr_pca

```

The next step imputes the data.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# To clean data
set.seed(123)
imputed <- impute_rf(compressed)

```

We can inspect PCA plot after imputation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
imp_bp <- plot_sample_boxplots(imputed,
                               order_by = "Group",
                               fill_by = "Group")
# PCA
imp_pca <- plot_pca(imputed,
                    center = TRUE,
                    shape = "Group",
                    color = "Group")
# Plot
imp_pca + imp_bp

```

Finally the data is ready to be exported and proceed with the statistical analysis.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

save(imputed, file = paste0(ppath, "Result/LCMS/Notame_LCMS_output.RData"))

```

# PCA plots

Droping flagged features

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Extract clean data
no_flag <- drop_flagged(imputed)

# Extracting feature height table
peak_height <- exprs(no_flag)
# Extracting samples information
pheno_data <- no_flag@phenoData@data
# Extracting feature information
feat_data <- no_flag@featureData@data

```

Preparing data and transposing feature table.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Transposing feature height table
transp_table  <- t(peak_height)
# Changing NA to 0 
transp_table[is.na(transp_table)]=0
# Centering and Scaling features
neg_pca <- prcomp(transp_table, center = TRUE, scale. = TRUE)

```

Plotting PCA results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Library to left_join use
library(dplyr)
# PCA scores
scores <- neg_pca$x %>%                  # Get PC coordinates
  data.frame %>%                         # Convert to data frames
  mutate(Sample_ID = rownames(.)) %>%    # Create a new column with the sample names
  left_join(pheno_data )                 # Adding metadata
# PCA plot
ggplot(scores,
       aes(PC1, PC2, shape = Species, color = Species)) +
  geom_point(size = 3) +
  guides(x=guide_axis(title = "PC1 (29.03 %)"),
         y=guide_axis(title = "PC2 (18.43 %)")) +
  theme_classic()
# Save plot
ggsave("Result/LCMS/LCMS_PCA.png", width = 5, height = 4, device='png', dpi="print")

```

Plotting loading results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

loadings <- neg_pca$rotation %>%           # Extract loadings
  data.frame(Cluster_ID = rownames(.))     # New column with feature name

```

Creating an artificial table with Feature name and Compound column.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Exporting notame output to find and filter identified metabolites
write_to_excel(clustered, "Result/LCMS/Clustered_features.xlsx")
# Load a metabolite name table
metab_name <- readxl::read_excel("Data/LCMS_Metabolites.xlsx", 1)
# Creating a new small table of the annotated compounds
neg_compouds <- left_join(metab_name, loadings)
# Plotting results
ggplot(loadings, aes(PC1, PC2)) + 
  geom_point(alpha = 0.2) +
  theme_classic() + 
  geom_point(data = neg_compouds,
             aes(shape = `Identification level`, color = `Identification level`),
             size = 2) +
  ggrepel::geom_label_repel(data = neg_compouds,
                            aes(label = Metabolite_name),
                            box.padding = 0.8,
                            label.padding = 0.3,
                            label.r = 0.3,
                            cex = 3) +
  guides(x=guide_axis(title = "PC1 (29.03 %)"),
         y=guide_axis(title = "PC2 (18.43 %)")) +
  ggsci::scale_color_aaas()
# Save plot
ggsave('Result/LCMS/LCMS_Loadings.png', width = 10, height = 6, device='png', dpi="print")

```

# Heat map plot

ComplexHeatmap package and dependency installation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# ComplexHeatmap package installation
#if (!requireNamespace("BiocManager", quietly=TRUE))
#    install.packages("BiocManager")
#BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)

# ColorRamp2 package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("jokergoo/colorRamp2")
library(colorRamp2)

# Cowplot package installation
#install.packages("cowplot")
library(cowplot)

```

Extracting and loaded of identified metabolites abundance.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Exporting notame output to find and filter height of identified metabolites
write_to_excel(no_flag, "Result/LCMS/no_flag_features.xlsx")
# Metabolite name table
metab_name_hm <- readxl::read_excel("Data/LCMS_Metabolite_HeatMap.xlsx", 2)
# Logarithmic scale
hm_scl <- metab_name_hm[, 6:9] %>% as.matrix %>% log10()
rownames(hm_scl) <- metab_name_hm$Metabolite_name
# Metabolite classification
metab_class <- metab_name_hm %>% select(Superclass, Metabolite = Metabolite_name)
# Metabolite class to HeatMap anotation
met_class_annotation <-  metab_class %>% select(Superclass) %>% 
  as.matrix()
rownames(met_class_annotation) <- metab_class$Metabolite
# Top information
top_info <- data.frame(Species = c( rep("A. rigidum"),
                                    rep("C. guianensis"),
                                    rep("M. laevis"),
                                    rep("P. sagotianum"))) 
rownames(top_info) <- paste(top_info$Species, rep(c(1, 2, 3, 4)))
top_info <- as.matrix(top_info)

```

Scaling, row and top heatmap anotation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

set.seed(2024)
# Metabolite class color
cols_metclass <- c("Benzenoids" = "#800000FF",
                   "Organic oxygen compounds" = "#8A9045FF",
                   "Lipids and lipid-like molecules" = "#FFA319FF",
                   "Phenylpropanoids and polyketides" = "#FF5A5A",
                   "Organic acids and derivatives" = "#155F83FF")
# Add row anotation to HeatMap
hm_row_ann <- rowAnnotation(Metabolite = met_class_annotation,
                            col = list(Metabolite = cols_metclass),
                            show_annotation_name = T,
                            show_legend=F)
# Species color
cols_species <- c("A. rigidum" = "#e6550d",
                 "C. guianensis" = "#91cf60",
                 "M. laevis" = "#33d2ff",
                 "P. sagotianum" = "#d957f8")
# Add top anotation to HeatMap
top_info_ann <- HeatmapAnnotation(`Species` = top_info,
                                  col = list(`Species` = cols_species),
                                  show_annotation_name = T,
                                  show_legend=F, 
                                  border = TRUE)
# Color scale
mycol <- colorRamp2(c(3, 5, 7),
                    c("blue", "white", "red"))
# Heatmap matrix plotting
hm_plot <- Heatmap(hm_scl,
        col = mycol,
        border_gp = grid::gpar(col = "black", lty = 0.05),
        rect_gp = grid::gpar(col = "black", lwd = 0.75),
        clustering_distance_columns = "euclidean",
        clustering_method_columns = "complete",
        top_annotation = top_info_ann,
        right_annotation = hm_row_ann,
        show_heatmap_legend = F,
        row_km = 3, column_km = 2)
hm_plot

```

Adding legends to heatmap.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Color scale legend
lgd1 <- Legend(col_fun = mycol,
               title = "log10 abundance",
               at = seq(7),
               direction = "horizontal" )
# Plants species legend
lgd2 <- Legend(labels = gt_render(c("*A. rigidum*",
                                    "*C. guianensis*",
                                    "*M. laevis*",
                                    "*P. sagotianum*")),
               legend_gp = gpar(fill = cols_species),
               title = "Plant species", ncol = 1)
# Metabolite class Legend
lgd3 <- Legend(labels = c(unique(metab_class$Superclass)),
               legend_gp = gpar(fill = cols_metclass), 
               title = "Metabolite superclass", ncol = 2)

```

ComplexHeatmap plot

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

set.seed(2024)
# Converting to ggplot
gg_heatmap <- grid.grabExpr(draw(hm_plot))
gg_heatmap <- ggpubr::as_ggplot(gg_heatmap)
# Legends
all_legends <- packLegend(lgd1, lgd2, lgd3, direction = "horizontal")
gg_legend <- grid.grabExpr(draw(all_legends))
gg_legend_fn <- ggpubr::as_ggplot(gg_legend)
# Heatmap plot
gcms_hm <- plot_grid(gg_legend_fn,
          gg_heatmap, ncol = 1,
          rel_heights = c(0.857, 4))
gcms_hm
# Save heatmap plot
ggsave(filename = "Result/LCMS/LCMS_Heatmap.jpeg", plot = gcms_hm,
      width = 5, height = 4, units = "in", dpi = 300, scale = 1.7)

```

Finish a record.

```{r}

finish_log()

```



