---
title: "Untargeted characterization of Amazonian barks extracts plants by Liquid and Gas Chromatography - Mass Spectrometry | Liquid Chromatography - Mass Spectrometry (LC-MS) data analysis 2"
author: "Jefferson Pastuña"
date: "2023-10-20"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

## Introduction

This R Script aims to record the procedure given for metabolic profiling of 4 species of plant used in Amazonia folk medicine. Each step has a brief explanation, code and graphics.

The workflow used was taken from ["notame": Workflow for Non-Targeted LC–MS Metabolic Profiling](https://doi.org/10.3390/metabo10040135). Which offers a wide variety of functions to perform metabolomic profile analysis.

## Before to start

The "notame" package accepts as input a feature table that can be obtained through software such as MZmine, MS-DIAL, among others. In this case, the feature table was obtained with the help of MZmine. The (*.csv) file was fixed to obtain the final feature table input.

## Notame workflow

As a first step for the analysis,"notame" package and other dependency packages were installed.

```{r echo=TRUE, message=FALSE}

# Notame package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("antonvsdata/notame")

# Notame library call
library(notame)

# Dependency packages installation
install_dependencies

```

Then, a main path and a log system was added to have a record of each process executed.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Main path
ppath <- "F:/Jefferson-Pastuna/Arajuno_Stem_Bark/"
# Log system
init_log(log_file = paste0(ppath, "Result/LCMS/Second_LCMS_Dataset_log.txt"))

```

Next, the MZmine suitable feature list was imported.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

data <- read_from_excel(file = "Data/Second_LCMS_Dataset_to_R.xlsx", sheet = 3, 
                        corner_row = 5, corner_column = "I", 
                        split_by = c("Column", "Ion Mode"))

```

Once the data is read, the next step was to create a MetaboSet in order to obtain a specific R object.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

modes <- construct_metabosets(exprs = data$exprs, 
                              pheno_data = data$pheno_data, 
                              feature_data = data$feature_data,
                              group_col = "Group")

```

We can visualize the raw data in order to inspect the processing routines.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Data extraction
mode_test <- modes$RP_NEG
# Boxplot of raw data
raw_bp <- plot_sample_boxplots(mode_test,
                               order_by = "Group",
                               fill_by = "Group")
# PCA of raw data
raw_pca <- plot_pca(mode_test,
                       center = TRUE,
                       shape = "Group",
                       color = "Group")
# Package to plots visualization in a same windows
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("thomasp85/patchwork")
library(patchwork)
# Plot
raw_pca + raw_bp

```

## Preprocessing

The first step of the preprocessing is to change the features with value equal to 0 to NA.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Data extraction
mode <- modes$RP_NEG
# Change 0 value to NA
mode <- mark_nas(mode, value = 0)

```

Then, features with low detection rate are first flagged and then could be removed. The "notame" package employs two criteria to select this features. First, is the feature presence in a percentage of QC injections, and then the feature presence in a percentage within a sample group or class.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Low detection rate
mode <- flag_detection(mode, qc_limit = 3/4, group_limit = 2/3)

```

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Some statistics after low detection algorithm
visualizations(mode, prefix = paste0(ppath, "Figure/LCMS/", "2nd_Low_Det"))

```

With these values, features which that were not detected in the 75% of the QC injections and 67% of sample groups will be flagged as low detection rate.

The next step for preprocessing correspond to drift correction. The drift correction can be applied by smoothed cubic spline regression.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drift correction
corrected <- correct_drift(mode)
# Flag low quality features
corrected <- flag_quality(corrected)

```

Then we can visualize the data after drift correction.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
corr_bp <- plot_sample_boxplots(corrected,
                                      order_by = "Group",
                                      fill_by = "Group")
# PCA
corr_pca <- plot_pca(corrected,
                        center = TRUE,
                        shape = "Group",
                        color = "Group") 
# Plot
corr_pca + corr_bp

```

Contaminant peaks based on the process blank will be removed.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Removal of contaminants
corrected_no_blank <- flag_contaminants(corrected,
                                        blank_col = "Group",
                                        blank_label = "PB",
                                        flag_thresh = 0.80,
                                        flag_label = "Contaminant")
# Removal blank group from dataset
corrected_no_blank <- corrected_no_blank[, corrected_no_blank$Group != "PB"]

```

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Some statistics after low detection algorithm
visualizations(corrected_no_blank, prefix = paste0(ppath, "Figure/LCMS/", "2nd_No_Blank"))

```

The next step removes the QC from the analysis, since they will not be needed in subsequent treatments.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

corrected_no_qc <- drop_qcs(corrected_no_blank)

```

We can visualize data without QC.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
no_qc_bp <- plot_sample_boxplots(corrected_no_qc,
                                 order_by = "Group",
                                 fill_by = "Group")
# PCA
no_qc_pca <- plot_pca(corrected_no_qc,
                      center = TRUE,
                      shape = "Group",
                      color = "Group")
# Plot
no_qc_pca + no_qc_bp

```

The next step is feature clustering. This step helps us reduce the number of features of the same molecule that were split due to ionization behavior (In-source fragmentation for example).

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

clustered <- cluster_features(corrected_no_qc,
                              rt_window = 1/60,
                              all_features = TRUE,
                              corr_thresh = 0.90,
                              d_thresh = 0.8
                              #plotting = TRUE,
                              #prefix = paste0(ppath, "Cluster/LCMS/H2OMeOH_Cluster")
                              )
compressed <- compress_clusters(clustered)

```

We can inspect PCA plot after clustering algorithm.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
compr_bp <- plot_sample_boxplots(compressed,
                                 order_by = "Group",
                                 fill_by = "Group")
# PCA
compr_pca <- plot_pca(compressed,
                      center = TRUE,
                      shape = "Group",
                      color = "Group")
# Plot
compr_pca + compr_bp

```

The next step imputes the data.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# To clean data
set.seed(123)
imputed <- impute_rf(compressed)

```

We can inspect PCA plot after imputation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
imp_bp <- plot_sample_boxplots(imputed,
                               order_by = "Group",
                               fill_by = "Group")
# PCA
imp_pca <- plot_pca(imputed,
                    center = TRUE,
                    shape = "Group",
                    color = "Group")
# Plot
imp_pca + imp_bp

```

Finally, the data is ready to be exported and proceed with the statistical analysis.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

save(imputed, file = paste0(ppath, "Result/LCMS/2nd_LCMS_notame_output.RData"))

```

# PCA plots

Droping flagged features

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Extract clean data
no_flag <- drop_flagged(imputed)
# Extracting feature height table
peak_height <- exprs(no_flag)
# Extracting samples information
pheno_data <- no_flag@phenoData@data
# Extracting feature information
feat_data <- no_flag@featureData@data

```

Preparing data and transposing feature table.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Transposing feature height table
transp_table  <- t(peak_height)
# Changing NA to 0 
transp_table[is.na(transp_table)]=0
# Centering and Scaling features
neg_pca <- prcomp(transp_table, center = TRUE, scale. = TRUE)

```

Plotting PCA results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Library to left_join use
library(dplyr)
# PCA scores
scores <- neg_pca$x %>%                   # Get PC coordinates
  data.frame %>%                          # Convert to data frames
  mutate(Sample_ID = rownames(.)) %>%     # Create a new column with the sample names
  left_join(pheno_data )                  # Adding metadata
# PCA plot
ggplot(scores,
       aes(PC1, PC2, shape = Species, color = Species)) +
  geom_point(size = 3) +
  guides(x=guide_axis(title = "PC1 (21.78 %)"),
         y=guide_axis(title = "PC2 (15.55 %)")) +
  theme_classic() +
  theme(legend.text = element_text(face="italic"))+
  theme(legend.position = c(0.87, 0.21),
        legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray")
# Save plot
ggsave("Result/LCMS/2nd_LCMS_PCA_to_DQ.png", width = 6, height = 4, device='png', dpi="print")

```

Plotting loading results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

loadings <- neg_pca$rotation %>%           # Extract loadings
  data.frame(Cluster_ID = rownames(.))     # New column with feat name

```

Creating an artificial table with feature ID and compound name column.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Load a metabolite name table
metab_name <- readxl::read_excel("Data/Second_LCMS_Metabolites.xlsx", 1)
# Creating a new table of identifield metabolites
compouds_tab <- left_join(metab_name, loadings)
# Plotting results
ggplot(loadings, aes(PC1, PC2)) + 
  geom_point(alpha = 0.2) +
  theme_classic() + 
  geom_point(data = compouds_tab, size = 1) +
  ggrepel::geom_label_repel(data = compouds_tab,
                            aes(label = Metabolite_name),
                            box.padding = 0.8,
                            label.padding = 0.3,
                            label.r = 0.3,
                            cex = 3) +
  guides(x=guide_axis(title = "PC1 (21.78 %)"),
         y=guide_axis(title = "PC2 (15.55 %)")) +
  ggsci::scale_color_aaas()
# Save plot
ggsave('Result/LCMS/2nd_lCMS_Loadings.png', width = 7, height = 4, device='png', dpi="print")

```

Finish a record.

```{r}

finish_log()

```

