---
title: "Untargeted characterization of Amazonian barks extracts plants by Liquid and Gas Chromatography - Mass Spectrometry | Liquid Chromatography - Mass Spectrometry (LC-MS) data analysis 2"
author: "Jefferson Pastuña"
date: "2023-10-20"
output:
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
usethis::git_vaccinate()

```

## Introduction

This R Script aims to record the procedure given for metabolic profiling of 4 species of plant used in Amazonia folk medicine. Each step has a brief explanation, code and graphics.

The workflow used was taken from ["notame": Workflow for Non-Targeted LC–MS Metabolic Profiling](https://doi.org/10.3390/metabo10040135). Which offers a wide variety of functions to perform metabolomic profile analysis.

## Before to start

The "notame" package accepts as input a feature table that can be obtained through software such as MZmine, MS-DIAL, among others. In this case, the feature table was obtained with the help of MZmine. The (*.csv) file was fixed to obtain the final feature table input.

## Notame workflow

As a first step for the analysis,"notame" package and other dependency packages were installed.

```{r echo=TRUE, message=FALSE}

# Notame package installation
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("antonvsdata/notame")

# Notame library call
library(notame)

# Dependency packages installation
install_dependencies

```

Then, a main path and a log system was added to have a record of each process executed.

```{r echo=TRUE, message=TRUE, warning=FALSE, error=FALSE}

# Main path
ppath <- "F:/Jefferson-Pastuna/Arajuno_Stem_Bark/"
# Log system
init_log(log_file = paste0(ppath, "Result/LCMS/Second_LCMS_Dataset_log.txt"))

```

Next, the MZmine suitable feature list was imported.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

data2 <- read_from_excel(file = "Data/Second_LCMS_Dataset_to_R.xlsx", sheet = 3,
                         corner_row = 5, corner_column = "I",
                         split_by = c("Column", "Ion Mode"))

```

Once the data is read, the next step was to create a MetaboSet in order to obtain a specific R object.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

modes2 <- construct_metabosets(exprs = data2$exprs,
                               pheno_data = data2$pheno_data,
                               feature_data = data2$feature_data,
                               group_col = "Group")

```

We can visualize the raw data in order to inspect the processing routines.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Data extraction
mode2_test <- modes2$RP_NEG
# Boxplot of raw data
raw_bp2 <- plot_sample_boxplots(mode2_test,
                                order_by = "Group",
                                fill_by = "Group")
# PCA of raw data
raw_pca2 <- plot_pca(mode2_test,
                     center = TRUE,
                     shape = "Group",
                     color = "Group")
# Package to plots visualization in a same windows
#if (!requireNamespace("devtools", quietly = TRUE)) {
#  install.packages("devtools")
#}
#devtools::install_github("thomasp85/patchwork")
library(patchwork)
# Plot
raw_pca2 + raw_bp2

```

## Preprocessing

The first step of the preprocessing is to change the features with value equal to 0 to NA.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Data extraction
mode2 <- modes2$RP_NEG
# Change 0 value to NA
mode2 <- mark_nas(mode2, value = 0)

```

Then, features with low detection rate are first flagged and then could be removed. The "notame" package employs two criteria to select this features. First, is the feature presence in a percentage of QC injections, and then the feature presence in a percentage within a sample group or class.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Low detection rate
mode2 <- flag_detection(mode2, qc_limit = 3/4, group_limit = 2/3)

```

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Some statistics after low detection algorithm
#visualizations(mode2, prefix = paste0(ppath, "Figure/LCMS/", "2nd_Low_Det"))

```

With these values, features which that were not detected in the 75% of the QC injections and 67% of sample groups will be flagged as low detection rate.

The next step for preprocessing correspond to drift correction. The drift correction can be applied by smoothed cubic spline regression.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Drift correction
corrected2 <- correct_drift(mode2)
# Flag low quality features
corrected2 <- flag_quality(corrected2)

```

Then we can visualize the data after drift correction.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
corr_bp2 <- plot_sample_boxplots(corrected2,
                                 order_by = "Group",
                                 fill_by = "Group")
# PCA
corr_pca2 <- plot_pca(corrected2,
                      center = TRUE,
                      shape = "Group",
                      color = "Group") 
# Plot
corr_pca2 + corr_bp2

```

Contaminant peaks based on the process blank will be removed.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Removal of contaminants
corrected2_no_blank <- flag_contaminants(corrected2,
                                         blank_col = "Group",
                                         blank_label = "PB",
                                         flag_thresh = 0.80,
                                         flag_label = "Contaminant")
# Removal blank group from dataset
corrected2_no_blank <- corrected2_no_blank[, corrected2_no_blank$Group != "PB"]

```

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Some statistics after low detection algorithm
#visualizations(corrected2_no_blank, prefix = paste0(ppath, "Figure/LCMS/", "2nd_No_Blank"))

```

The next step removes the QC from the analysis, since they will not be needed in subsequent treatments.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

corrected2_no_qc <- drop_qcs(corrected2_no_blank)

```

We can visualize data without QC.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
no_qc_bp2 <- plot_sample_boxplots(corrected2_no_qc,
                                  order_by = "Group",
                                  fill_by = "Group")
# PCA
no_qc_pca2 <- plot_pca(corrected2_no_qc,
                       center = TRUE,
                       shape = "Group",
                       color = "Group")
# Plot
no_qc_pca2 + no_qc_bp2

```

The next step is feature clustering. This step helps us reduce the number of features of the same molecule that were split due to ionization behavior (In-source fragmentation for example).

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

clustered2 <- cluster_features(corrected2_no_qc,
                               rt_window = 1/60,
                               all_features = TRUE,
                               corr_thresh = 0.90,
                               d_thresh = 0.8
                               #plotting = TRUE,
                               #prefix = paste0(ppath, "Cluster/LCMS/H2OMeOH_Cluster")
                               )
compressed2 <- compress_clusters(clustered2)

```

We can inspect PCA plot after clustering algorithm.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
compr_bp2 <- plot_sample_boxplots(compressed2,
                                  order_by = "Group",
                                  fill_by = "Group")
# PCA
compr_pca2 <- plot_pca(compressed2,
                       center = TRUE,
                       shape = "Group",
                       color = "Group")
# Plot
compr_pca2 + compr_bp2

```

The next step imputes the data.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# To clean data
set.seed(123)
imputed2 <- impute_rf(compressed2)

```

We can inspect PCA plot after imputation.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

# Boxplot
imp_bp2 <- plot_sample_boxplots(imputed2,
                                order_by = "Group",
                                fill_by = "Group")
# PCA
imp_pca2 <- plot_pca(imputed2,
                     center = TRUE,
                     shape = "Group",
                     color = "Group")
# Plot
imp_pca2 + imp_bp2

```

Finally, the data is ready to be exported and proceed with the statistical analysis.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}

#save(imputed2, file = paste0(ppath, "Result/LCMS/2nd_LCMS_notame_output.RData"))

```

# PCA plots

Droping flagged features

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Extract clean data
no_flag2 <- drop_flagged(imputed2)
# Extracting feature height table
peak_height2 <- exprs(no_flag2)
# Extracting samples information
pheno_data2 <- no_flag2@phenoData@data
# Extracting feature information
feat_data2 <- no_flag2@featureData@data

```

Preparing data and transposing feature table.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Transposing feature height table
transp_table2  <- t(peak_height2)
# Changing NA to 0 
transp_table2[is.na(transp_table2)]=0
# Centering and Scaling features
neg_pca2 <- prcomp(transp_table2, center = TRUE, scale. = TRUE)

```

Plotting PCA results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Library to left_join use
library(dplyr)
# PCA scores
scores2 <- neg_pca2$x %>%                 # Get PC coordinates
  data.frame %>%                          # Convert to data frames
  mutate(Sample_ID = rownames(.)) %>%     # Create a new column with the sample names
  left_join(pheno_data2)                  # Adding metadata
# PCA plot
figure_1a <- ggplot(scores2,
                    aes(PC1, PC2, shape = Species, color = Species)) +
  geom_point(size = 3) +
  guides(x=guide_axis(title = "PC1 (21.78 %)"),
         y=guide_axis(title = "PC2 (15.55 %)")) +
  theme_classic() +
  theme(legend.text = element_text(face="italic"))+
  theme(legend.position = c(0.88, 0.50),
        legend.background = element_rect(fill = "white", color = "black")) +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  geom_vline(xintercept = 0, linetype = "longdash", colour="gray") +
  geom_hline(yintercept = 0, linetype = "longdash", colour="gray")
figure_1a
# Save plot
#ggsave("Result/LCMS/2nd_LCMS_PCA_to_DQ.png", width = 6, height = 4, device='png', dpi="print")

```

Plotting loading results.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

loadings2 <- neg_pca2$rotation %>%         # Extract loadings
  data.frame(Cluster_ID = rownames(.))     # New column with feat name

```

Creating an artificial table with feature ID and compound name column.

```{r echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, results='hide'}

# Load a metabolite name table
metab_name2 <- readxl::read_excel("Data/Second_LCMS_Metabolites.xlsx", 1)
# Creating a new table of identifield metabolites
compouds_tab2 <- left_join(metab_name2, loadings2)
# Plotting results
ggplot(loadings2, aes(PC1, PC2)) +
  geom_point(alpha = 0.2) +
  theme_classic() + 
  geom_point(data = compouds_tab2, size = 1) +
  ggrepel::geom_label_repel(data = compouds_tab2,
                            aes(label = Metabolite_name),
                            box.padding = 0.8,
                            label.padding = 0.3,
                            label.r = 0.3,
                            cex = 3) +
  guides(x=guide_axis(title = "PC1 (21.78 %)"),
         y=guide_axis(title = "PC2 (15.55 %)")) +
  ggsci::scale_color_aaas()
# Save plot
#ggsave('Result/LCMS/2nd_lCMS_Loadings.png', width = 7, height = 4, device='png', dpi="print")

```

Finish a record.

```{r}

finish_log()

```

